Verifying Spatial Properties of Stencil Computations

### 2016-11-15 paper.tex
Remove pagerefs and use Def. abbreviation consistently

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, Def.~\ref{def:neighbour},
  p.~\pageref{def:neighbour}) and $\infty$ is a wild-card
  representing both constant indices and un-specified indices. Throughout $u, v, w$ range over
  index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

@@

\paragraph{Core analysis}
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes.
This is captured by the function $\textit{analyse}$:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(a')\}}
\end{array}
\end{align*}
%
The predicate $\neigh$ classifies neighbourhood indices, based on
the induction variables of the loop (implicit here).
Thus, we focus on assignments to an array subscript where the LHS
index $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a \emph{multiset} of indexing schemes for its subscripts, computed
with \textsf{schematic} (Defn.~\ref{def:schematic},
p.~\pageref{def:schematic}).

@@

We illustrate each step of inference using the
 five-point stencil program from Example~\ref{exm:checking} (p.~\pageref{exm:checking}).

@@

\noindent
\textbf{2)\,} Sort each permutation set into an ordered list, by the ordering:
\begin{equation*}
  \vect{x} \leq \vect{y} =
      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
\end{equation*}
%
giving the sorting function $\textsf{sort}(M)$. For our example:
%%
\begin{align*}
\textsf{sort}(P(M_0)) =
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
\noindent
\textbf{3)\,} Reduce each list pairwise by $\bullet$ operation
(Defn.~\ref{def:span-coalesc}) to coalesce contiguous regions,
given by $\textsf{fold}\bullet$. For our example, this yields:
%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}
]), (\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\textbf{4)\,} Un-permute and union together, \ie{},
%
\[
U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
 \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
\]
For our example:
%
\begin{align*}
U(M) =
\{\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{-1}{0}{1},
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\textbf{5)\,} Filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller, defined:
%
\begin{align*}
& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\spanOp(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
Applying \spanOp{} again yields the same
result for our example; the fixed point is reached.

@@

Abstract syntax trees of the specification language
are synthesised from the set $\spanOp(M)$ of coverings spans.
This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}) and which is wrapped
in the $\textsf{Approx}$ type (see
Defn.~\ref{def:mult-and-approx},
p.~\pageref{def:mult-and-approx}).  The algebra mirrors the shape
of the region combinators for sum and product,
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
  \mathsf{Approx}(\specDNF) \\
\odot & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
        \mathsf{Approx}(\specDNF) \\
\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
\end{align*}
%%
The $\mathsf{toSpec}$ function maps a dimension identifier
and a pair of lower and upper bound
values (one-dimensional) to a specification with a one-dimensional region.
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) =
\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
We unpack the definitions of this algebra.

\begin{definition}[Specifications in DNF]
Let $r_1, ..., r_n$ range over elements of the \textit{region}
syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
The set \specDNF{} is a subset of syntax trees \textit{spec},
containing specifications in disjunctive normal form over regions
where the disjunct is \term{+} (since its model is in terms of set
union) and the conjunct is \term{*} (since its model is a form of tensor).
That is elements of \specDNF{} are of the form:
%%
$
(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots
\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
$.

@@

\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
  is defined in terms of the intermediate operator $\hat{\odot}$ on
  \specDNF{}, defined:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
 the (\trule{\textsc{dist}}) rule of our equational theory
 (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
commutative and associative, and sound with respect to our
model (see Lemma~\ref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
       \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u) \\
\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
                                              \hat{\odot} T)
\end{align*}
where in the last case $\textit{inj}$ corresponds to the unary
injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
as $\odot$ is commutative.
\end{definition}

>>>>

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, Def.~\ref{def:neighbour}) and $\infty$ is a wild-card
  representing both constant indices and un-specified indices. Throughout $u, v, w$ range over
  index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

@@

\paragraph{Core analysis}
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes.
This is captured by the function $\textit{analyse}$:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(a')\}}
\end{array}
\end{align*}
%
The predicate $\neigh$ classifies neighbourhood indices, based on
the induction variables of the loop (implicit here).
Thus, we focus on assignments to an array subscript where the LHS
index $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a \emph{multiset} of indexing schemes for its subscripts, computed
with \textsf{schematic} (Def.~\ref{def:schematic}).

@@

We illustrate each step of inference using the
 five-point stencil program from Example~\ref{exm:checking}.

@@

\noindent
\textbf{2)\,} Sort each permutation set into an ordered list, by the ordering:
\begin{equation*}
  \vect{x} \leq \vect{y} =
      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
\end{equation*}
%
giving the sorting function $\textsf{sort}(M)$. For our example:
%%
\begin{align*}
\textsf{sort}(P(M_0)) =
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
\noindent
\textbf{3)\,} Reduce each list pairwise by $\bullet$ operation
(Def.~\ref{def:span-coalesc}) to coalesce contiguous regions,
given by $\textsf{fold}\bullet$. For our example, this yields:
%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}
]), (\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\textbf{4)\,} Un-permute and union together, \ie{},
%
\[
U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
 \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
\]
For our example:
%
\begin{align*}
U(M) =
\{\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{-1}{0}{1},
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\textbf{5)\,} Filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller, defined:
%
\begin{align*}
& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\spanOp(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
Applying \spanOp{} again yields the same
result for our example; the fixed point is reached.

@@

Abstract syntax trees of the specification language
are synthesised from the set $\spanOp(M)$ of coverings spans.
This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}) and which is wrapped
in the $\textsf{Approx}$ type (see
Def.~\ref{def:mult-and-approx}).  The algebra mirrors the shape
of the region combinators for sum and product,
with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
%%
\begin{align*}
\oplus & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
  \mathsf{Approx}(\specDNF) \\
\odot & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
        \mathsf{Approx}(\specDNF) \\
\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
\end{align*}
%%
The $\mathsf{toSpec}$ function maps a dimension identifier
and a pair of lower and upper bound
values (one-dimensional) to a specification with a one-dimensional region.
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) =
\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
We unpack the definitions of this algebra.

\begin{definition}[Specifications in DNF]
Let $r_1, ..., r_n$ range over elements of the \textit{region}
syntax (see Figure~\ref{fig:syntax}).
The set \specDNF{} is a subset of syntax trees \textit{spec},
containing specifications in disjunctive normal form over regions
where the disjunct is \term{+} (since its model is in terms of set
union) and the conjunct is \term{*} (since its model is a form of tensor).
That is elements of \specDNF{} are of the form:
%%
$
(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots
\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
$.

@@

\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
  is defined in terms of the intermediate operator $\hat{\odot}$ on
  \specDNF{}, defined:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
 the (\trule{\textsc{dist}}) rule of our equational theory
 (Fig.~\ref{fig:equations}). The operation is
commutative and associative, and sound with respect to our
model (see Lemma~\ref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
       \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u) \\
\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
                                              \hat{\odot} T)
\end{align*}
where in the last case $\textit{inj}$ corresponds to the unary
injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
as $\odot$ is commutative.
\end{definition}
